A Parallel Algorithm for Solving Global Optimization Problems and Its Application for Tuning Hyperparameters of AI Methods

### 2025-06-17 2025/RuSCDays/RSD-2025.tex
minor correction
Add sentence about Optuna and HyperOpt

%(рус.) Таким образом, возникает the hyperparameter optimization (HPO) problem, которая с математической точки зрения является задачей глобальной оптимизации. Типичным является случай наличия не только непрерывных, но и категориальных параметров, т.е. возникает необходимость использования методы глобальной оптимизации, способные решать задачи со смешанными параметрами.
Thus, there arises the hyperparameter optimization (HPO) problem, which from the mathematical point of view is a global optimization problem. In this case, it is typical to have not only continuous but also categorical parameters (parameters having a prefixed set of values), i.e. there is a need to use global optimization methods capable of solving problems with mixed parameters.

@@

%(рус.) Пусть проведено k испытаний, их результаты получены и обработаны мастером, а на текущей итерации процессы-рабочие проводят испытания в точках множества S_k  = {x^(k+1),x^(k+2),…,x^(k+p) }, т.е. в этих точках испытания уже начались, но еще не завершены. Как только один из процессов-рабочих завершает проведение испытания в некоторой точке (например, x^(k+1)), он пересылает процессу-мастеру результаты испытания. Получив от рабочего процесса результаты испытания и обработав их мастер, в свою очередь, выбирает для него точку нового испытания x^(k+p+1) в соответствии со следующими правилами.
Suppose $k$ trials have been performed, the master has received and processed their results, and during the current iteration the worker processes perform trials at the points of the set $S_k={x^{k+1}, x^{k+2},..., x^{k+p}}$, i.e. the trials have already started at these points but have not yet been completed. As soon as one of the worker processes completes a trial at some point (e.g. $x^{k+1}$), it forwards the results of the trial to the master process. After receiving the trial results from the worker process and processing them, the master process in turn selects a new trial point $x^{k+p+1}$ for the worker process according to the following rules.

>>>>

%(рус.) Таким образом, возникает the hyperparameter optimization (HPO) problem, которая с математической точки зрения является задачей глобальной оптимизации. Типичным является случай наличия не только непрерывных, но и категориальных параметров, т.е. возникает необходимость использования методы глобальной оптимизации, способные решать задачи со смешанными параметрами.
Thus, there arises the hyperparameter optimization (HPO) problem, which from the mathematical point of view is a global optimization problem. In this case, it is typical to have not only continuous but also categorical parameters (parameters having a prefixed set of values), i.e. there is a need to use global optimization methods capable of solving problems with mixed parameters. 
Such methods belonging to the class of metaheuristic and model-based algorithms are implemented in the well-known hyperparameter optimization frameworks Optuna \cite{optuna} and HyperOpt \cite{hyperopt}.

@@

%(рус.) Пусть проведено k испытаний, их результаты получены и обработаны мастером, а на текущей итерации процессы-рабочие проводят испытания в точках множества S_k  = {x^(k+1),x^(k+2),…,x^(k+p) }, т.е. в этих точках испытания уже начались, но еще не завершены. Как только один из процессов-рабочих завершает проведение испытания в некоторой точке (например, x^(k+1)), он пересылает процессу-мастеру результаты испытания. Получив от рабочего процесса результаты испытания и обработав их мастер, в свою очередь, выбирает для него точку нового испытания x^(k+p+1) в соответствии со следующими правилами.
Suppose $k$ trials have been performed, the master has received and processed their results, and during the current iteration the worker processes perform trials at the points of the set $S_k={x^{k+1}, x^{k+2},..., x^{k+p}}$, i.e. the trials have already started at these points but have not yet been completed. As soon as one of the worker processes completes a trial at some point (e.g. $x^{k+1}$), it forwards the results of the trial to the master process and the point is excluded from the set $S_k$. After receiving the trial results from the worker process and processing them, the master process in turn selects a new trial point $x^{k+p+1}$ for the worker process according to the following rules.
